---
title: "CST Typing by DESI-MS - Comparison between different Community State Types"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir='./')
```

Load the required packages
```{r, include=FALSE}
library(ggplot2)
library(doParallel)
library(readr)
library(reshape2)
library(caret)
library(plotROC)
library(pROC)
library(precrec)
set.seed(648791)
# Set ggplot titles to appear at center
ggplot() + theme(plot.title = element_text(hjust = 'center'))
registerDoParallel(cores=22)
dir.create('./CST_Comparisons_DESI-MS/', recursive=TRUE)
```

# Data import
Load the VMET2 DESI-MS Positive and Negative mode datasets
```{r, warning=F, echo=F, message=F}
# Load the datasets
desi_neg_vmet2 <- read.table("../../Data/VMET2_DESI-MS_NEG.csv", header=1, sep=',')
desi_pos_vmet2 <- read.table("../../Data/VMET2_DESI-MS_POS.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet2 <- read_csv("../../Data/VMET2_Metadata.csv")
cst_metadata_vmet2 <- read_csv("../../Data/VMET2_CSTAssignment.csv")

study_metadata_vmet2 <- merge(study_metadata_vmet2, cst_metadata_vmet2, by='Seq_ID')

# Merge with the clinical metadata
desi_pos_vmet2 <- merge(study_metadata_vmet2, desi_pos_vmet2, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
desi_neg_vmet2 <- merge(study_metadata_vmet2, desi_neg_vmet2, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
colnames(desi_pos_vmet2)[colnames(desi_pos_vmet2) == 'Seq_ID.x'] <- 'Seq_ID'
colnames(desi_neg_vmet2)[colnames(desi_neg_vmet2) == 'Seq_ID.x'] <- 'Seq_ID'

# Load the data QC results and exclude samples with low spectral quality
desi_neg_vmet2_QC <- read.table("./DESI-MS QC/VMET2_DESI_Negative_QC.csv", header=1, sep=',')
desi_pos_vmet2_QC <- read.table("./DESI-MS QC/VMET2_DESI_Positive_QC.csv", header=1, sep=',')

# Exclude the samples flagged in the DESI-MS Data QC Notebook
neg_vmet2_exclusions <- desi_neg_vmet2[desi_neg_vmet2_QC$Exclude == TRUE, 'DESI_ID']
pos_vmet2_exclusions <- desi_pos_vmet2[desi_pos_vmet2_QC$Exclude == TRUE, 'DESI_ID']
desi_neg_vmet2 <- desi_neg_vmet2[!(desi_neg_vmet2$DESI_ID %in% neg_vmet2_exclusions), ]
desi_pos_vmet2 <- desi_pos_vmet2[!(desi_pos_vmet2$DESI_ID %in% pos_vmet2_exclusions), ]

rm(study_metadata_vmet2)
rm(cst_metadata_vmet2)
rm(desi_neg_vmet2_QC)
rm(desi_pos_vmet2_QC)
rm(neg_vmet2_exclusions)
rm(pos_vmet2_exclusions)
```

Load the VMET DESI-MS Positive and Negative mode datasets
```{r, warning=F, echo=F, message=F}
# Load the DESI-MS datasets
desi_neg_vmet <- read.table("../../Data/VMET_DESI-MS_NEG.csv", header=1, sep=',')
desi_pos_vmet <- read.table("../../Data/VMET_DESI-MS_POS.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet <- read_csv("../../Data/VMET_Metadata.csv")
cst_metadata_vmet <- read_csv("../../Data/VMET_CSTAssignment.csv")

study_metadata_vmet <- merge(study_metadata_vmet, cst_metadata_vmet, by='Seq_ID')

# Merge with the clinical metadata
desi_pos_vmet <- merge(study_metadata_vmet, desi_pos_vmet, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
desi_neg_vmet <- merge(study_metadata_vmet, desi_neg_vmet, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
colnames(desi_pos_vmet)[colnames(desi_pos_vmet) == 'Seq_ID.x'] <- 'Seq_ID'
colnames(desi_neg_vmet)[colnames(desi_neg_vmet) == 'Seq_ID.x'] <- 'Seq_ID'

# Load the data QC results and exclude samples with low spectral quality
desi_neg_vmet_QC <- read.table("./DESI-MS QC/VMET_DESI_Negative_QC.csv", header=1, sep=',')
desi_pos_vmet_QC <- read.table("./DESI-MS QC/VMET_DESI_Positive_QC.csv", header=1, sep=',')

# Exclude the samples flagged in the DESI-MS Data QC Notebook
neg_vmet_exclusions <- desi_neg_vmet[desi_neg_vmet_QC$Exclude == TRUE, 'DESI_ID']
pos_vmet_exclusions <- desi_pos_vmet[desi_pos_vmet_QC$Exclude == TRUE, 'DESI_ID']
desi_neg_vmet <- desi_neg_vmet[!(desi_neg_vmet$DESI_ID %in% neg_vmet_exclusions), ]
desi_pos_vmet <- desi_pos_vmet[!(desi_pos_vmet$DESI_ID %in% pos_vmet_exclusions), ]

rm(study_metadata_vmet)
rm(cst_metadata_vmet)
rm(desi_neg_vmet_QC)
rm(desi_pos_vmet_QC)
rm(neg_vmet_exclusions)
rm(pos_vmet_exclusions)
```

Overview of dataset dimensions
```{r}
# VMET2 DESI-MS Negative Mode
cat(paste("VMET2 DESI-MS Negative Mode Matrix: ", dim(desi_neg_vmet2)[1], "samples, ", dim(desi_neg_vmet2)[2] - 29, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desineg_first_metabolite_idx_vmet2 <- 29

# VMET2 DESI-MS Positive Mode
cat(paste("VMET2 DESI-MS Positive Mode Matrix: ", dim(desi_pos_vmet2)[1], "samples, ", dim(desi_pos_vmet2)[2] - 29, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desipos_first_metabolite_idx_vmet2 <- 29

# VMET DESI-MS Negative Mode
cat(paste("VMET DESI-MS Negative Mode Matrix: ", dim(desi_neg_vmet)[1], "samples, ", dim(desi_neg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desineg_first_metabolite_idx_vmet <- 28

# VMET DESI-MS Positive Mode
cat(paste("VMET DESI-MS Positive Mode Matrix: ", dim(desi_pos_vmet)[1], "samples, ", dim(desi_pos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desipos_first_metabolite_idx_vmet <- 28
```
# Discrimination of CST types from each other based on their DESI-MS metabolic profiles
In this analysis we examine if the DESI-MS profiles are significantly differenty and can be to discriminate between individual Community State Types (CST). A series of functions defined in the next cell will handle the model fitting and cross-validation for each specific comparison (IvsIII, IvsIV, IIIvsV, IIIvsIV, IvsV). 
The interface from the *caret* package is used to fit the random forest models and perform repeated (n=15 times) 7-fold stratified cross validation.

```{r}
# I vs III
random_forest_comparison_IvsIII <- function(dataset, first_metabo_idx) {
  dataset_train <- dataset[, c('CST', colnames(dataset)[first_metabo_idx:dim(dataset)[2]])]
  dataset_train$CST <- factor(dataset_train$CST)
  dataset_train[is.na(dataset_train)] <- 0
  
  dataset_train <- subset(dataset_train, CST == 'I' | CST == 'III')
  
  dataset_train$CST <- droplevels(dataset_train$CST)
  dataset_train$CST <- relevel(dataset_train$CST, 'I')
  
  fit_control <- trainControl(method = "repeatedcv", repeats=15, classProbs = TRUE,
                           number = 7, summaryFunction=multiClassSummary, preProcOptions = c('center', 'scale'), savePredictions=TRUE)
  rf_fit <- train(CST ~ ., 
                   data = dataset_train, method='rf', trControl=fit_control, ntree=1000, metric='AUC', tuneGrid=data.frame(mtry=dim(dataset_train)[2]/3))
  
  return (rf_fit)

}

# I vs IV
random_forest_comparison_IvsIV <- function(dataset, first_metabo_idx) {
  dataset_train <- dataset[, c('CST', colnames(dataset)[first_metabo_idx:dim(dataset)[2]])]
  dataset_train$CST <- factor(dataset_train$CST)

  dataset_train[is.na(dataset_train)] <- 0
  
  dataset_train <- subset(dataset_train, CST == 'I' | CST == 'IV')
  
  dataset_train$CST <- droplevels(dataset_train$CST)
  dataset_train$CST <- relevel(dataset_train$CST, 'I')
   
  fit_control <- trainControl(method = "repeatedcv", repeats=15, classProbs = TRUE,
                           number = 7, summaryFunction=multiClassSummary, 
                           preProcOptions = c('center', 'scale'), savePredictions=TRUE)
  rf_fit <- train(CST ~ ., 
                 data = dataset_train, method='rf', trControl=fit_control, ntree=1000, metric='AUC', 
                 tuneGrid=data.frame(mtry=dim(dataset_train)[2]/3))

  
  return (rf_fit)
}

# III vs IV
random_forest_comparison_IIIvsIV <- function(dataset, first_metabo_idx) {
  dataset_train <- dataset[, c('CST', colnames(dataset)[first_metabo_idx:dim(dataset)[2]])]
  dataset_train$CST <- factor(dataset_train$CST)

  dataset_train[is.na(dataset_train)] <- 0
  
  dataset_train <- subset(dataset_train, CST == 'III' | CST == 'IV')
  
  dataset_train$CST <- droplevels(dataset_train$CST)
  dataset_train$CST <- relevel(dataset_train$CST, 'III')

  fit_control <- trainControl(method = "repeatedcv", repeats=15, classProbs = TRUE,
                           number = 7, summaryFunction=multiClassSummary, 
                           preProcOptions = c('center', 'scale'), savePredictions=TRUE)
  rf_fit <- train(CST ~ ., 
                   data = dataset_train, method='rf', trControl=fit_control, ntree=1000, metric='AUC',
                  tuneGrid=data.frame(mtry=dim(dataset_train)[2]/3))

  return (rf_fit)
}

# III vs V
random_forest_comparison_IIIvsV <- function(dataset, first_metabo_idx) {
  dataset_train <- dataset[, c('CST', colnames(dataset)[first_metabo_idx:dim(dataset)[2]])]
  dataset_train$CST <- factor(dataset_train$CST)

  dataset_train[is.na(dataset_train)] <- 0
  
  dataset_train <- subset(dataset_train, CST == 'III' | CST == 'V')
  
  dataset_train$CST <- droplevels(dataset_train$CST)
  dataset_train$CST <- relevel(dataset_train$CST, 'III')

  fit_control <- trainControl(method = "repeatedcv", repeats=15, classProbs = TRUE,
                           number = 7, summaryFunction=multiClassSummary, 
                           preProcOptions = c('center', 'scale'), savePredictions=TRUE)
  rf_fit <- train(CST ~ ., 
                              data = dataset_train, method='rf', trControl=fit_control, ntree=1000, metric='AUC',
                  tuneGrid=data.frame(mtry=dim(dataset_train)[2]/3))


  
  return (rf_fit)
}

# I vs V
# Applies downsampling of CST I samples to match the number of CST IVs. All CST IV samples
# are always used, but in each round of CV different CST III samples are sampled
random_forest_comparison_IvsV <- function(dataset, first_metabo_idx) {
  dataset_train <- dataset[, c('CST', colnames(dataset)[first_metabo_idx:dim(dataset)[2]])]
  dataset_train$CST <- factor(dataset_train$CST)

  dataset_train[is.na(dataset_train)] <- 0
  
  dataset_train <- subset(dataset_train, CST == 'I' | CST == 'V')
  
  dataset_train$CST <- droplevels(dataset_train$CST)
  dataset_train$CST <- relevel(dataset_train$CST, 'V')

  fit_control <- trainControl(method = "repeatedcv", repeats=15, classProbs = TRUE,
                           number = 7, summaryFunction=multiClassSummary, 
                           preProcOptions = c('center', 'scale'), savePredictions=TRUE)
  rf_fit <- train(CST ~ ., 
                              data = dataset_train, method='rf', trControl=fit_control, ntree=1000, metric='AUC',
                  tuneGrid=data.frame(mtry=dim(dataset_train)[2]/3))

  return (rf_fit)
}

# Helper function to calculate parameters for roc curves and generate quick plot
get_roc <- function(model, levels, target) {
  
    #cv_original_class <- model$trainingData$.outcome
    cv_original_class <- model$pred$obs
    #cv_predictor_avg <- aggregate(as.formula(paste(target, "~ rowIndex")),model$pred,mean)[,target]
    cv_predictor_avg <- model$pred[, target]
    roc_test_set <- roc(cv_original_class, cv_predictor_avg, ci=TRUE)
    roc_data <- data.frame(Class=cv_original_class,target=cv_predictor_avg)
    roc_plot <- ggplot(roc_data, 
    aes(m = target, d = factor(Class, levels = c(levels)), color='firebrick')) + guides(color = FALSE) + 
    geom_roc(hjust = 0, vjust = 0) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') +  annotate("label", x = .75, y = .25, 
           label = paste("Mean AUC =", round(roc_test_set$auc, 2)), color='firebrick')
    
    return(list(data=roc_data, roc=roc_test_set, plot=roc_plot)) 
}

```

# Metabolic Differences between Community State Types
## Discrimination between CST I vs III
Lactobacillus crispatus (CST I) versus Lactobacillus iners (CST III).
```{r}
# Fit models to VMET2 
IvsIII_desipos_vmet2 <- random_forest_comparison_IvsIII(desi_pos_vmet2,  desipos_first_metabolite_idx_vmet2)
IvsIII_desineg_vmet2 <- random_forest_comparison_IvsIII(desi_neg_vmet2, desineg_first_metabolite_idx_vmet2)
# Fit models to VMET
IvsIII_desipos_vmet <- random_forest_comparison_IvsIII(desi_pos_vmet,  desipos_first_metabolite_idx_vmet)
IvsIII_desineg_vmet <- random_forest_comparison_IvsIII(desi_neg_vmet, desineg_first_metabolite_idx_vmet)


# save the model performance to .txt files
capture.output(print(IvsIII_desipos_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_POS_RF_Model_IvsIII.txt')
capture.output(print(IvsIII_desipos_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_POS_RF_Model_IvsIII.txt')
capture.output(print(IvsIII_desineg_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_NEG_RF_Model_IvsIII.txt')
capture.output(print(IvsIII_desineg_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_NEG_RF_Model_IvsIII.txt')

# ROC_Information
desipos_roc_vmet2 <- get_roc(IvsIII_desipos_vmet2, c('I', 'III'), 'III')
desineg_roc_vmet2 <- get_roc(IvsIII_desineg_vmet2, c('I', 'III'), 'III')
desipos_roc_vmet <- get_roc(IvsIII_desipos_vmet, c('I', 'III'), 'III')
desineg_roc_vmet <- get_roc(IvsIII_desineg_vmet, c('I', 'III'), 'III')

# Stack the roc curves on a dataframe for plotting
pos_roc_data_vmet2 <- cbind(desipos_roc_vmet2$data, Mode=rep('Positive', dim(desipos_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desipos_roc_vmet2$data)[1]))
neg_roc_data_vmet2 <- cbind(desineg_roc_vmet2$data, Mode=rep('Negative', dim(desineg_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desineg_roc_vmet2$data)[1]))

pos_roc_data_vmet <- cbind(desipos_roc_vmet$data, Mode=rep('Positive', dim(desipos_roc_vmet$data)[1]), Study=rep('VMET', dim(desipos_roc_vmet$data)[1]))
neg_roc_data_vmet <- cbind(desineg_roc_vmet$data, Mode=rep('Negative', dim(desineg_roc_vmet$data)[1]), Study=rep('VMET', dim(desineg_roc_vmet$data)[1]))

desi_roc_data <- rbind(pos_roc_data_vmet2, neg_roc_data_vmet2, pos_roc_data_vmet, neg_roc_data_vmet)

rm(pos_roc_data_vmet2)
rm(neg_roc_data_vmet2)
rm(pos_roc_data_vmet)
rm(neg_roc_data_vmet)

desi_roc_data$Class <- factor(desi_roc_data$Class, levels = c('I', 'III'))
desi_roc_data$Class <- as.numeric(desi_roc_data$Class) - 1

#roc_lab <- paste(paste("Negative mode: Mean AUC =", round(desineg_roc$roc$auc, 2)), '\n', paste("Positive mode: Mean AUC =", round(desipos_roc$roc$auc, 2)), sep='')

roc_Ldepleted_desi <- ggplot(desi_roc_data, 
    aes(m = target, d = Class, color=Study, linetype=Mode)) + 
    geom_roc(hjust = 0, vjust = 0, labels=F) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') + ggtitle('CST I vs CST III - DESI-MS') + scale_linetype_manual(values=c("dotted", "solid"))#+ annotate("label", x = c(.75), y = c(.25), parse=FALSE, label = roc_lab)

ggsave(filename='./CST_Comparisons_DESI-MS/DESI-MS_IvsIII_CST_ROC.png', roc_Ldepleted_desi)
```

## Discrimination between CST I vs IV
Lactobacillus crispatus (CST I) versus CST IV
```{r}
# Fit models to VMET2 
IvsIV_desipos_vmet2 <- random_forest_comparison_IvsIV(desi_pos_vmet2,  desipos_first_metabolite_idx_vmet2)
IvsIV_desineg_vmet2 <- random_forest_comparison_IvsIV(desi_neg_vmet2, desineg_first_metabolite_idx_vmet2)
# Fit models to VMET
IvsIV_desipos_vmet <- random_forest_comparison_IvsIV(desi_pos_vmet,  desipos_first_metabolite_idx_vmet)
IvsIV_desineg_vmet <- random_forest_comparison_IvsIV(desi_neg_vmet, desineg_first_metabolite_idx_vmet)


# save the model performance to .txt files
capture.output(print(IvsIV_desipos_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_POS_RF_Model_IvsIV.txt')
capture.output(print(IvsIV_desipos_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_POS_RF_Model_IvsIV.txt')
capture.output(print(IvsIV_desineg_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_NEG_RF_Model_IvsIV.txt')
capture.output(print(IvsIV_desineg_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_NEG_RF_Model_IvsIV.txt')

# ROC_Information
desipos_roc_vmet2 <- get_roc(IvsIV_desipos_vmet2, c('I', 'IV'), 'IV')
desineg_roc_vmet2 <- get_roc(IvsIV_desineg_vmet2, c('I', 'IV'), 'IV')
desipos_roc_vmet <- get_roc(IvsIV_desipos_vmet, c('I', 'IV'), 'IV')
desineg_roc_vmet <- get_roc(IvsIV_desineg_vmet, c('I', 'IV'), 'IV')

# Stack the roc curves on a dataframe for plotting
pos_roc_data_vmet2 <- cbind(desipos_roc_vmet2$data, Mode=rep('Positive', dim(desipos_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desipos_roc_vmet2$data)[1]))
neg_roc_data_vmet2 <- cbind(desineg_roc_vmet2$data, Mode=rep('Negative', dim(desineg_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desineg_roc_vmet2$data)[1]))

pos_roc_data_vmet <- cbind(desipos_roc_vmet$data, Mode=rep('Positive', dim(desipos_roc_vmet$data)[1]), Study=rep('VMET', dim(desipos_roc_vmet$data)[1]))
neg_roc_data_vmet <- cbind(desineg_roc_vmet$data, Mode=rep('Negative', dim(desineg_roc_vmet$data)[1]), Study=rep('VMET', dim(desineg_roc_vmet$data)[1]))

desi_roc_data <- rbind(pos_roc_data_vmet2, neg_roc_data_vmet2, pos_roc_data_vmet, neg_roc_data_vmet)

rm(pos_roc_data_vmet2)
rm(neg_roc_data_vmet2)
rm(pos_roc_data_vmet)
rm(neg_roc_data_vmet)

desi_roc_data$Class <- factor(desi_roc_data$Class, levels = c('I', 'IV'))
desi_roc_data$Class <- as.numeric(desi_roc_data$Class) - 1

#roc_lab <- paste(paste("Negative mode: Mean AUC =", round(desineg_roc$roc$auc, 2)), '\n', paste("Positive mode: Mean AUC =", round(desipos_roc$roc$auc, 2)), sep='')

roc_Ldepleted_desi <- ggplot(desi_roc_data, 
    aes(m = target, d = Class, color=Study, linetype=Mode)) + 
    geom_roc(hjust = 0, vjust = 0, labels=F) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') + ggtitle('CST I vs CST IV - DESI-MS') + scale_linetype_manual(values=c("dotted", "solid"))#+ annotate("label", x = c(.75), y = c(.25), parse=FALSE, label = roc_lab)

ggsave(filename='./CST_Comparisons_DESI-MS/DESI-MS_IvsIV_CST_ROC.png', roc_Ldepleted_desi)
```

## Discrimination between CST III vs IV
Lactobacillus iners (CST III) versus CST IV
```{r}
# Fit models to VMET2 
IIIvsIV_desipos_vmet2 <- random_forest_comparison_IIIvsIV(desi_pos_vmet2,  desipos_first_metabolite_idx_vmet2)
IIIvsIV_desineg_vmet2 <- random_forest_comparison_IIIvsIV(desi_neg_vmet2, desineg_first_metabolite_idx_vmet2)
# Fit models to VMET
IIIvsIV_desipos_vmet <- random_forest_comparison_IIIvsIV(desi_pos_vmet,  desipos_first_metabolite_idx_vmet)
IIIvsIV_desineg_vmet <- random_forest_comparison_IIIvsIV(desi_neg_vmet, desineg_first_metabolite_idx_vmet)

# save the model performance to .txt files
capture.output(print(IIIvsIV_desipos_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_POS_RF_Model_IIIvsIV.txt')
capture.output(print(IIIvsIV_desipos_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_POS_RF_Model_IIIvsIV.txt')
capture.output(print(IIIvsIV_desineg_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_NEG_RF_Model_IIIvsIV.txt')
capture.output(print(IIIvsIV_desineg_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_NEG_RF_Model_IIIvsIV.txt')

# ROC_Information
desipos_roc_vmet2 <- get_roc(IIIvsIV_desipos_vmet2, c('III', 'IV'), 'IV')
desineg_roc_vmet2 <- get_roc(IIIvsIV_desineg_vmet2, c('III', 'IV'), 'IV')
desipos_roc_vmet <- get_roc(IIIvsIV_desipos_vmet, c('III', 'IV'), 'IV')
desineg_roc_vmet <- get_roc(IIIvsIV_desineg_vmet, c('III', 'IV'), 'IV')

# Stack the roc curves on a dataframe for plotting
pos_roc_data_vmet2 <- cbind(desipos_roc_vmet2$data, Mode=rep('Positive', dim(desipos_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desipos_roc_vmet2$data)[1]))
neg_roc_data_vmet2 <- cbind(desineg_roc_vmet2$data, Mode=rep('Negative', dim(desineg_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desineg_roc_vmet2$data)[1]))

pos_roc_data_vmet <- cbind(desipos_roc_vmet$data, Mode=rep('Positive', dim(desipos_roc_vmet$data)[1]), Study=rep('VMET', dim(desipos_roc_vmet$data)[1]))
neg_roc_data_vmet <- cbind(desineg_roc_vmet$data, Mode=rep('Negative', dim(desineg_roc_vmet$data)[1]), Study=rep('VMET', dim(desineg_roc_vmet$data)[1]))

desi_roc_data <- rbind(pos_roc_data_vmet2, neg_roc_data_vmet2, pos_roc_data_vmet, neg_roc_data_vmet)

rm(pos_roc_data_vmet2)
rm(neg_roc_data_vmet2)
rm(pos_roc_data_vmet)
rm(neg_roc_data_vmet)

desi_roc_data$Class <- factor(desi_roc_data$Class, levels = c('III', 'IV'))
desi_roc_data$Class <- as.numeric(desi_roc_data$Class) - 1

#roc_lab <- paste(paste("Negative mode: Mean AUC =", round(desineg_roc$roc$auc, 2)), '\n', paste("Positive mode: Mean AUC =", round(desipos_roc$roc$auc, 2)), sep='')

roc_Ldepleted_desi <- ggplot(desi_roc_data, 
    aes(m = target, d = Class, color=Study, linetype=Mode)) + 
    geom_roc(hjust = 0, vjust = 0, labels=F) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') + ggtitle('CST III vs CST IV - DESI-MS') + scale_linetype_manual(values=c("dotted", "solid"))#+ annotate("label", x = c(.75), y = c(.25), parse=FALSE, label = roc_lab)

ggsave(filename='./CST_Comparisons_DESI-MS/DESI-MS_IIIvsIV_CST_ROC.png', roc_Ldepleted_desi)
```


## Discrimination between CST III vs IV
Lactobacillus iners (CST III) versus Lactobacillus crispatus (CST I)
```{r}
# Fit models to VMET2 
IIIvsV_desipos_vmet2 <- random_forest_comparison_IIIvsV(desi_pos_vmet2,  desipos_first_metabolite_idx_vmet2)
IIIvsV_desineg_vmet2 <- random_forest_comparison_IIIvsV(desi_neg_vmet2, desineg_first_metabolite_idx_vmet2)
# Fit models to VMET
IIIvsV_desipos_vmet <- random_forest_comparison_IIIvsV(desi_pos_vmet,  desipos_first_metabolite_idx_vmet)
IIIvsV_desineg_vmet <- random_forest_comparison_IIIvsV(desi_neg_vmet, desineg_first_metabolite_idx_vmet)

# save the model performance to .txt files
capture.output(print(IIIvsV_desipos_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_POS_RF_Model_IIIvsV.txt')
capture.output(print(IIIvsV_desipos_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_POS_RF_Model_IIIvsV.txt')
capture.output(print(IIIvsV_desineg_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_NEG_RF_Model_IIIvsV.txt')
capture.output(print(IIIvsV_desineg_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_NEG_RF_Model_IIIvsV.txt')

# ROC_Information
desipos_roc_vmet2 <- get_roc(IIIvsV_desipos_vmet2, c('III', 'V'), 'V')
desineg_roc_vmet2 <- get_roc(IIIvsV_desineg_vmet2, c('III', 'V'), 'V')
desipos_roc_vmet <- get_roc(IIIvsV_desipos_vmet, c('III', 'V'), 'V')
desineg_roc_vmet <- get_roc(IIIvsV_desineg_vmet, c('III', 'V'), 'V')

# Stack the roc curves on a dataframe for plotting
pos_roc_data_vmet2 <- cbind(desipos_roc_vmet2$data, Mode=rep('Positive', dim(desipos_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desipos_roc_vmet2$data)[1]))
neg_roc_data_vmet2 <- cbind(desineg_roc_vmet2$data, Mode=rep('Negative', dim(desineg_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desineg_roc_vmet2$data)[1]))

pos_roc_data_vmet <- cbind(desipos_roc_vmet$data, Mode=rep('Positive', dim(desipos_roc_vmet$data)[1]), Study=rep('VMET', dim(desipos_roc_vmet$data)[1]))
neg_roc_data_vmet <- cbind(desineg_roc_vmet$data, Mode=rep('Negative', dim(desineg_roc_vmet$data)[1]), Study=rep('VMET', dim(desineg_roc_vmet$data)[1]))

desi_roc_data <- rbind(pos_roc_data_vmet2, neg_roc_data_vmet2, pos_roc_data_vmet, neg_roc_data_vmet)

rm(pos_roc_data_vmet2)
rm(neg_roc_data_vmet2)
rm(pos_roc_data_vmet)
rm(neg_roc_data_vmet)

desi_roc_data$Class <- factor(desi_roc_data$Class, levels = c('III', 'V'))
desi_roc_data$Class <- as.numeric(desi_roc_data$Class) - 1

#roc_lab <- paste(paste("Negative mode: Mean AUC =", round(desineg_roc$roc$auc, 2)), '\n', paste("Positive mode: Mean AUC =", round(desipos_roc$roc$auc, 2)), sep='')

roc_Ldepleted_desi <- ggplot(desi_roc_data, 
    aes(m = target, d = Class, color=Study, linetype=Mode)) + 
    geom_roc(hjust = 0, vjust = 0, labels=F) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') + ggtitle('CST III vs CST V - DESI-MS') + scale_linetype_manual(values=c("dotted", "solid"))#+ annotate("label", x = c(.75), y = c(.25), parse=FALSE, label = roc_lab)

ggsave(filename='./CST_Comparisons_DESI-MS/DESI-MS_IIIvsV_CST_ROC.png', roc_Ldepleted_desi)
```

## Discrimination between CST I vs V
Lactobacillus crispatus (CST I) versus Lactobacillus jensenii (CST V)
```{r}
# Fit models to VMET2 
IvsV_desipos_vmet2 <- random_forest_comparison_IvsV(desi_pos_vmet2,  desipos_first_metabolite_idx_vmet2)
IvsV_desineg_vmet2 <- random_forest_comparison_IvsV(desi_neg_vmet2, desineg_first_metabolite_idx_vmet2)
# Fit models to VMET
IvsV_desipos_vmet <- random_forest_comparison_IvsV(desi_pos_vmet,  desipos_first_metabolite_idx_vmet)
IvsV_desineg_vmet <- random_forest_comparison_IvsV(desi_neg_vmet, desineg_first_metabolite_idx_vmet)


# save the model performance to .txt files
capture.output(print(IvsV_desipos_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_POS_RF_Model_IvsV.txt')
capture.output(print(IvsV_desipos_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_POS_RF_Model_IvsV.txt')
capture.output(print(IvsV_desineg_vmet2), file='./CST_Comparisons_DESI-MS/VMET2_DESI_NEG_RF_Model_IvsV.txt')
capture.output(print(IvsV_desineg_vmet), file='./CST_Comparisons_DESI-MS/VMET_DESI_NEG_RF_Model_IvsV.txt')

# ROC_Information
desipos_roc_vmet2 <- get_roc(IvsV_desipos_vmet2, c('I', 'V'), 'V')
desineg_roc_vmet2 <- get_roc(IvsV_desineg_vmet2, c('I', 'V'), 'V')
desipos_roc_vmet <- get_roc(IvsV_desipos_vmet, c('I', 'V'), 'V')
desineg_roc_vmet <- get_roc(IvsV_desineg_vmet, c('I', 'V'), 'V')

# Stack the roc curves on a dataframe for plotting
pos_roc_data_vmet2 <- cbind(desipos_roc_vmet2$data, Mode=rep('Positive', dim(desipos_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desipos_roc_vmet2$data)[1]))
neg_roc_data_vmet2 <- cbind(desineg_roc_vmet2$data, Mode=rep('Negative', dim(desineg_roc_vmet2$data)[1]), Study=rep('VMET2', dim(desineg_roc_vmet2$data)[1]))

pos_roc_data_vmet <- cbind(desipos_roc_vmet$data, Mode=rep('Positive', dim(desipos_roc_vmet$data)[1]), Study=rep('VMET', dim(desipos_roc_vmet$data)[1]))
neg_roc_data_vmet <- cbind(desineg_roc_vmet$data, Mode=rep('Negative', dim(desineg_roc_vmet$data)[1]), Study=rep('VMET', dim(desineg_roc_vmet$data)[1]))

desi_roc_data <- rbind(pos_roc_data_vmet2, neg_roc_data_vmet2, pos_roc_data_vmet, neg_roc_data_vmet)

rm(pos_roc_data_vmet2)
rm(neg_roc_data_vmet2)
rm(pos_roc_data_vmet)
rm(neg_roc_data_vmet)

desi_roc_data$Class <- factor(desi_roc_data$Class, levels = c('I', 'V'))
desi_roc_data$Class <- as.numeric(desi_roc_data$Class) - 1

#roc_lab <- paste(paste("Negative mode: Mean AUC =", round(desineg_roc$roc$auc, 2)), '\n', paste("Positive mode: Mean AUC =", round(desipos_roc$roc$auc, 2)), sep='')

roc_Ldepleted_desi <- ggplot(desi_roc_data, 
    aes(m = target, d = Class, color=Study, linetype=Mode)) + 
    geom_roc(hjust = 0, vjust = 0, labels=F) + coord_equal() + style_roc(theme=theme_grey, xlab='1 - Specificity', ylab='Sensitivity') + ggtitle('CST I vs CST V - DESI-MS') + scale_linetype_manual(values=c("dotted", "solid"))#+ annotate("label", x = c(.75), y = c(.25), parse=FALSE, label = roc_lab)

ggsave(filename='./CST_Comparisons_DESI-MS/DESI-MS_IvsV_CST_ROC.png', roc_Ldepleted_desi)
```
