---
title: "CST Typing by DESI-MS - R2 Linear mixed effect model analysis"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir='./')
```

Load the required packages
```{r, include=FALSE}
library(lme4)
library(ggplot2)
library(doParallel)
library(r2glmm)
library(readr)
library(reshape2)
library(MuMIn)
library(emmeans)

set.seed(774321)
# Set ggplot titles to appear at center
ggplot() + theme(plot.title = element_text(hjust = 'center'))
registerDoParallel(cores=22)
dir.create('./LinearMixedModels_LC-MS/', recursive=TRUE)
dir.create('./LinearMixedModels_LC-MS/rSquared', recursive=TRUE)

```

# Data import
Load the VMET LC-MS datasets
```{r, warning=F, echo=F, message=F}
# Load the DESI-MS datasets
lc_lneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_LNEG.csv", header=1, sep=',')
lc_lpos_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_LPOS.csv", header=1, sep=',')
lc_rneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_RNEG.csv", header=1, sep=',')
lc_rpos_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_RPOS.csv", header=1, sep=',')
lc_hneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_HNEG.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet <- read_csv("../../Data/VMET_Metadata.csv")
cst_metadata_vmet <- read_csv("../../Data/VMET_CSTAssignment.csv")

study_metadata_vmet <- merge(study_metadata_vmet, cst_metadata_vmet, by='Seq_ID')

# Merge with the clinical metadata
lc_lneg_vmet <- merge(study_metadata_vmet, lc_lneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_lpos_vmet <- merge(study_metadata_vmet, lc_lpos_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_rneg_vmet <- merge(study_metadata_vmet, lc_rneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_rpos_vmet <- merge(study_metadata_vmet, lc_rpos_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_hneg_vmet <- merge(study_metadata_vmet, lc_hneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)

rm(study_metadata_vmet)
rm(cst_metadata_vmet)
```

Overview of dataset dimensions
```{r}
# VMET LC-MS Lipid Negative Mode
cat(paste("VMET LC-MS Lipid Negative Mode Matrix: ", dim(lc_lneg_vmet)[1], "samples, ", dim(lc_lneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_lneg_first_metabolite_idx_vmet <- 28

# VMET LC-MS Lipid Positive Mode
cat(paste("VMET LC-MS Lipid Positive Mode Matrix: ", dim(lc_lpos_vmet)[1], "samples, ", dim(lc_lpos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_lpos_first_metabolite_idx_vmet <- 28

# VMET LC-MS Reversed Phase Negative Mode
cat(paste("VMET LC-MS Reversed Phase Negative Mode Matrix: ", dim(lc_rneg_vmet)[1], "samples, ", dim(lc_rneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_rneg_first_metabolite_idx_vmet <- 28

# VMET LC-MS Reversed Phase Positive Mode
cat(paste("VMET LC-MS Reversed Phase Positive Mode Matrix: ", dim(lc_rpos_vmet)[1], "samples, ", dim(lc_rpos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_rpos_first_metabolite_idx_vmet <- 28

# VMET LC-MS HILIC Negative Mode
cat(paste("VMET LC-MS HILIC Negative Mode Matrix: ", dim(lc_hneg_vmet)[1], "samples, ", dim(lc_hneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_hneg_first_metabolite_idx_vmet <- 28
```

# Main Linear mixed effect model structure and contrasts
In this notebook, we will define a general mixed effect model formula to model metabolite variability as function of Community State Type, (CST), gestational age, maternal BMI, maternal age, ethnicity, with random effects (slope with gestational age and intercept for baseline differences) to account for between individual variability.

Example LME Model definition and summary
```{r}
# Select a random feature from VMET2
currmet <- colnames(desi_neg_vmet2)[1085]

# metabolite variability explained as a function of CST, GestationalAge, 
# BMI and Ethnicity, with random intercept and slope (GestationalAge) per patient
rsmodel <- lmer(log(get(currmet) + 1) ~ CST + GestationalAge +  BMI + Ethnicity + Age + (0 + GestationalAge|SubjectID) + (1|SubjectID), data=desi_neg_vmet2,  REML=TRUE)

# Default model summmary
summary(rsmodel, ddf='Kenward-Roger')
```

## Linear mixed effect modelling of the VMET2 Dataset
In the next cell we define an utility function which fits a linear mixed model for a single metabolic feature and retrieves the main estimates and p-values. This function will be applied in parallel to all variables in a DESI-MS dataset. This version of the function is tailored for the VMET2 cohort metadata naming conventions. Model is adjusted for CST, GestationalAge, BMI, Ethnicity and Age, with a random slope and intercept per individual. Contrasts for differences between CSTs, including LDominant (average of I, II, III, V) vs LDepleted CST (IV) are calculated for mid-Gestation (20 weeks), with maternal age = 30, BMI=25 and averaged over all Ethnicity levels. Comparisons between early and late gestation (36 vs 8 weeks) are also calculated, with maternal age=30, BMI=25 and averaged over all Community State Types and Ethnicity levels.

Define another helper function to profile the partial R2 per covariate in parallel
```{r}
fit_lmm_r2estimates <- function(dataset, var) {
    tryCatch({
      
      # Get metabolite name
      currmet <- colnames(dataset)[var]
      
      model_variables <- c(currmet, 'CST', 'GestationalAge', 'Age', 'BMI', 'Ethnicity', 'SubjectID')
      complete_idx <- complete.cases(dataset[model_variables])
      
      model_formula <- as.formula(paste0(currmet, " ~ CST + GestationalAge + Age +  BMI + Ethnicity + (GestationalAge||SubjectID)"))
      currentDataset <- dataset[complete_idx, model_variables]
      currentDataset[, 1] <- log(currentDataset[, 1] + 1)
      rsmodel <- lmer(model_formula, data=currentDataset,  REML=TRUE)
      
      r2_partial <- data.frame(r2beta(rsmodel, method='kr'))
      
      rownames(r2_partial) <- r2_partial$Effect
      # Re-order rows so the row binding is coherent in foreach
      r2_partial <- r2_partial[c('Model', 'CST', 'GestationalAge', 'BMI', 'Ethnicity', 'Age'), ]
      mumin_r2 <- r.squaredGLMM(rsmodel)
      r2_cond <- mumin_r2[1]
      r2_marg <- mumin_r2[2]
      icc <- performance::icc(rsmodel, tolerance = 1e-999)
      
      r2_values <- as.numeric(c(r2_partial$Rsq, r2_cond, r2_marg, icc$ICC_adjusted))
      r2_names <- c(as.character(r2_partial$Effect), 'R2Conditional_nsj', 'R2Marginal_nsj', 'Adjusted_ICC')
      r2_names[1] <- 'R2_Model_kr'
      
      pval_values <- as.numeric(c(r2_partial$pval, NA, NA, NA))
      pval_names <- lapply(r2_names, FUN=function(x) {paste0(x, '_pval')})
      
      res_frame <- data.frame(cbind(substring(currmet, first=2), t(r2_values), t(pval_values)))
      colnames(res_frame) <- c("Metabolite", r2_names, pval_names)
      
      return(res_frame)
      
    }, error = function(err){stop()})
}
```

Iterate through the VMET DESI Positive ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_neg_vmet <- foreach(var=desineg_first_metabolite_idx_vmet:dim(desi_neg_vmet)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_neg_vmet, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_neg_vmet)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_neg_vmet[2:dim(r2_estimates_desi_neg_vmet)[2]] <- apply(r2_estimates_desi_neg_vmet[, 2:dim(r2_estimates_desi_neg_vmet)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(r2_estimates_desi_neg_vmet, './LinearMixedModels_LC-MS/rSquared/VMET_DESI_NEG_LME_R2.csv')
```

Iterate through the VMET DESI Negative ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_pos_vmet <- foreach(var=desipos_first_metabolite_idx_vmet:dim(desi_pos_vmet)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_pos_vmet, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_pos_vmet)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_pos_vmet[2:dim(r2_estimates_desi_pos_vmet)[2]] <- apply(r2_estimates_desi_pos_vmet[, 2:dim(r2_estimates_desi_pos_vmet)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)


write.csv(r2_estimates_desi_pos_vmet, './LinearMixedModels_LC-MS/rSquared/VMET_DESI_POS_LME_R2.csv')
```

Iterate through the VMET 2 DESI Negative ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_neg_vmet2 <- foreach(var=desineg_first_metabolite_idx_vmet2:dim(desi_neg_vmet2)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_neg_vmet2, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_neg_vmet2)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_neg_vmet2[2:dim(r2_estimates_desi_neg_vmet2)[2]] <- apply(r2_estimates_desi_neg_vmet2[, 2:dim(r2_estimates_desi_neg_vmet2)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(r2_estimates_desi_neg_vmet2, './LinearMixedModels_LC-MS/rSquared/VMET2_DESI_NEG_LME_R2.csv')
```

Iterate through the VMET 2 DESI Positive ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_pos_vmet2 <- foreach(var=desipos_first_metabolite_idx_vmet2:dim(desi_pos_vmet2)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_pos_vmet2, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_pos_vmet2)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_pos_vmet2[2:dim(r2_estimates_desi_pos_vmet2)[2]] <- apply(r2_estimates_desi_pos_vmet2[, 2:dim(r2_estimates_desi_pos_vmet2)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)


write.csv(r2_estimates_desi_pos_vmet2, './LinearMixedModels_LC-MS/rSquared/VMET2_DESI_POS_LME_R2.csv')
```

# Generate a stacked barplot barplot with the R2 measures
```{r}
plotR2 <- function(r2_results, top_n) {
  
  r2_values_set <- r2_results[, c("Metabolite", "CST", "GestationalAge", "Age", "BMI", "Ethnicity", "R2Conditional_nsj", "R2_Model_kr")]
  r2_values_set[, 9] <- r2_values_set[, 7] - rowSums(r2_values_set[, 2:6])
  r2_values_set[r2_values_set[, 9] < 0, 9] <- 0

  colnames(r2_values_set)[9] <- 'R2'

  melted <- melt(r2_values_set, 'Metabolite')

  melted$Metabolite <- round(as.numeric(as.character(melted$Metabolite)), digits=3)
  melted$value <- as.numeric(melted$value)

  covar_to_sort_by <- 'R2Conditional_nsj'
  r2_covar <- subset(melted, variable == covar_to_sort_by)
  r2_covar <- r2_covar[order(-r2_covar$value), ]
  melted$Metabolite <- round(as.numeric(as.character(melted$Metabolite)), digits=3)
  melted$variable <- factor(melted$variable, levels = c('R2Conditional_nsj', 'R2', 'CST', 'GestationalAge', 'Ethnicity', 'BMI', 'Age'))
  top_n_mets <- r2_covar[1:top_n, 1]

  plotDFrame <- melted[melted$Metabolite %in% top_n_mets, ]
  plotDFrame <- subset(plotDFrame, variable != 'R2Conditional_nsj')
  
  r2_plot <- ggplot(plotDFrame, aes(fill=variable, y=as.numeric(value), x=factor(Metabolite, r2_covar$Metabolite[1:top_n]))) +  geom_bar(stat='identity', position='stack', alpha=1) + ylab('R2') + xlab('Metabolite') + theme( axis.text.x = element_blank(),
  axis.ticks.x = element_blank())
  return (r2_plot)
}
```

# Plots for top 100 features (ranked by variance explained in linear model, including the SubjectID) 
```{r}
vmet2_neg_top100 <- plotR2(r2_estimates_desi_neg_vmet2, 100)
vmet_neg_top100 <- plotR2(r2_estimates_desi_neg_vmet, 100)
vmet_pos_top100 <- plotR2(r2_estimates_desi_pos_vmet, 100)
vmet2_pos_top100 <- plotR2(r2_estimates_desi_pos_vmet2, 100)

ggsave('./rSquared/VMET2_DESINEG_100R2.png', vmet2_neg_top100)
ggsave('./rSquared/VMET_DESINEG_100R2.png', vmet_neg_top100)
ggsave('./rSquared/VMET2_DESIPOS_100R2.png', vmet2_pos_top100)
ggsave('./rSquared/VMET_DESIPOS_100R2.png', vmet_pos_top100)
```

# Variance explained for the Markers shown in Figure 1 (B) - main replicated hits of linear mixed modeling analysis.
```{r}
library(xlsx)
ldepl_markers <- read.xlsx('../../Data/Extended Data Table 2 LDOM vs LDEPL.xlsx', sheetIndex=1)

ldepl_markers_neg <- subset(ldepl_markers, DESI.MS.mode =='negative')

ldepl_markers_pos <- subset(ldepl_markers, DESI.MS.mode =='positive')

ldepl_vmet_neg <- r2_estimates_desi_neg_vmet[r2_estimates_desi_neg_vmet$Metabolite %in% ldepl_markers_neg$VMET_m.z.measured, ]
ldepl_vmet2_neg <- r2_estimates_desi_neg_vmet2[r2_estimates_desi_neg_vmet2$Metabolite %in% ldepl_markers_neg$VMET2_m.z.measured, ]

ldepl_vmet_pos <- r2_estimates_desi_pos_vmet[r2_estimates_desi_pos_vmet$Metabolite %in% ldepl_markers_pos$VMET_m.z.measured, ]
ldepl_vmet2_pos <- r2_estimates_desi_pos_vmet2[r2_estimates_desi_pos_vmet2$Metabolite %in% ldepl_markers_pos$VMET2_m.z.measured, ]

formatMz <- function(x) {sapply(x, FUN=function(y){format(round(y, 4), nsmall = 4)})}

immune_vmet2_neg <- r2_estimates_desi_neg_vmet2[formatMz(as.numeric(r2_estimates_desi_neg_vmet2$Metabolite)) %in% formatMz(immune_markers$m.z.measured), ]

vmet_neg_topMarkers <- plotR2(ldepl_vmet_neg, 150)
vmet2_neg_topMarkers <- plotR2(ldepl_vmet2_neg, 150)

vmet_pos_topMarkers <- plotR2(ldepl_vmet_pos, 150)
vmet2_pos_topMarkers <- plotR2(ldepl_vmet2_pos, 150)

ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET_DESINEG_R2_LDomLDeplMarkers.png', vmet_neg_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET2_DESINEG_R2_LDomLDeplMarkers.png', vmet2_neg_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET2_DESIPOS_R2_LDomLDeplMarkers.png', vmet2_pos_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET_DESIPOS_R2_LDomLDeplMarkers.png', vmet_pos_topMarkers)
```

# Multivariate analysis of variance - PERMANOVA
```{r}
library(vegan)
library(gt)
library(dplyr)

# VMET HNEG
metabolite_matrix <- lc_hneg_vmet[, lc_hneg_first_metabolite_idx_vmet:dim(lc_hneg_vmet)[2]]
permanovaDataset <- lc_hneg_vmet[, c("CST", "GestationalAge", "Ethnicity" , "BMI", "Age", "SubjectID")]
metabolite_matrix <- metabolite_matrix[complete.cases(permanovaDataset), ]
permanovaDataset  <- permanovaDataset[complete.cases(permanovaDataset), ]
permanova_vmet_hneg <- adonis2(log(metabolite_matrix + 1) ~ CST + GestationalAge + Ethnicity + BMI + Age + SubjectID, data = permanovaDataset, method='euclidean')

results_table <-
  permanova_vmet_hneg %>%
  gt(rownames_to_stub = TRUE) %>%
  fmt_number(
    columns = c(4,5,6),
    decimals = 3,
    suffixing = F
  ) %>%
  fmt_missing(columns = c(5,6), 
              missing_text = '')
gtsave(results_table, './LinearMixedModels_LC-MS/rSquared/VMET_HNEG_PERMANOVA.html')


# VMET LNEG
metabolite_matrix <- lc_lneg_vmet[, lc_lneg_first_metabolite_idx_vmet:dim(lc_lneg_vmet)[2]]
permanovaDataset <- lc_lneg_vmet[, c("CST", "GestationalAge", "Ethnicity" , "BMI", "Age", "SubjectID")]
metabolite_matrix <- metabolite_matrix[complete.cases(permanovaDataset), ]
permanovaDataset  <- permanovaDataset[complete.cases(permanovaDataset), ]
permanova_vmet_lneg <- adonis2(log(metabolite_matrix + 1) ~ CST + GestationalAge + Ethnicity + BMI + Age + SubjectID, data = permanovaDataset, method='euclidean')

results_table <-
  permanova_vmet_lneg %>%
  gt(rownames_to_stub = TRUE) %>%
  fmt_number(
    columns = c(4,5,6),
    decimals = 3,
    suffixing = F
  ) %>%
  fmt_missing(columns = c(5,6), 
              missing_text = '')
gtsave(results_table, './LinearMixedModels_LC-MS/rSquared/VMET_LNEG_PERMANOVA.html')

# VMET RNEG
metabolite_matrix <- lc_rneg_vmet[, lc_rneg_first_metabolite_idx_vmet:dim(lc_rneg_vmet)[2]]
permanovaDataset <- lc_rneg_vmet[, c("CST", "GestationalAge", "Ethnicity" , "BMI", "Age", "SubjectID")]
metabolite_matrix <- metabolite_matrix[complete.cases(permanovaDataset), ]
permanovaDataset  <- permanovaDataset[complete.cases(permanovaDataset), ]
permanova_vmet_rneg <- adonis2(log(metabolite_matrix + 1) ~ CST + GestationalAge + Ethnicity + BMI + Age + SubjectID, data = permanovaDataset, method='euclidean')

results_table <-
  permanova_vmet_rneg %>%
  gt(rownames_to_stub = TRUE) %>%
  fmt_number(
    columns = c(4,5,6),
    decimals = 3,
    suffixing = F
  ) %>%
  fmt_missing(columns = c(5,6), 
              missing_text = '')
gtsave(results_table, './LinearMixedModels_LC-MS/rSquared/VMET_RNEG_PERMANOVA.html')


# VMET LPOS
metabolite_matrix <- lc_lpos_vmet[, lc_lpos_first_metabolite_idx_vmet:dim(lc_lpos_vmet)[2]]
permanovaDataset <- lc_lpos_vmet[, c("CST", "GestationalAge", "Ethnicity" , "BMI", "Age", "SubjectID")]
metabolite_matrix <- metabolite_matrix[complete.cases(permanovaDataset), ]
permanovaDataset  <- permanovaDataset[complete.cases(permanovaDataset), ]
permanova_vmet_lpos <- adonis2(log(metabolite_matrix + 1) ~ CST + GestationalAge + Ethnicity + BMI + Age + SubjectID, data = permanovaDataset, method='euclidean')

results_table <-
  permanova_vmet_lpos %>%
  gt(rownames_to_stub = TRUE) %>%
  fmt_number(
    columns = c(4,5,6),
    decimals = 3,
    suffixing = F
  ) %>%
  fmt_missing(columns = c(5,6), 
              missing_text = '')
gtsave(results_table, './LinearMixedModels_LC-MS/rSquared/VMET_LPOS_PERMANOVA.html')


# VMET RPOS
metabolite_matrix <- lc_rpos_vmet[, lc_rpos_first_metabolite_idx_vmet:dim(lc_rpos_vmet)[2]]
permanovaDataset <- lc_rpos_vmet[, c("CST", "GestationalAge", "Ethnicity" , "BMI", "Age", "SubjectID")]
metabolite_matrix <- metabolite_matrix[complete.cases(permanovaDataset), ]
permanovaDataset  <- permanovaDataset[complete.cases(permanovaDataset), ]
permanova_vmet_rpos <- adonis2(log(metabolite_matrix + 1) ~ CST + GestationalAge + Ethnicity + BMI + Age + SubjectID, data = permanovaDataset, method='euclidean')

results_table <-
  permanova_vmet_rpos %>%
  gt(rownames_to_stub = TRUE) %>%
  fmt_number(
    columns = c(4,5,6),
    decimals = 3,
    suffixing = F
  ) %>%
  fmt_missing(columns = c(5,6), 
              missing_text = '')
gtsave(results_table, './LinearMixedModels_LC-MS/rSquared/VMET_RPOS_PERMANOVA.html')
```