---
title: "CST Typing by LC-MS - Detection of Lactobacillus Depleted Status"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir='./')
```

Load the required packages
```{r, include=FALSE}
library(lme4)
library(ggplot2)
library(sjPlot)
library(doParallel)
library(r2glmm)
library(readr)
library(reshape2)
library(MuMIn)
library(emmeans)
library(pROC)

set.seed(3046802)
# Set ggplot titles to appear at center
ggplot() + theme(plot.title = element_text(hjust = 'center'))
registerDoParallel(cores=22)
dir.create('./LinearMixedModels_LC-MS/', recursive=TRUE)

```

# Data import
Load the VMET LC-MS datasets
```{r, warning=F, echo=F, message=F}
# Load the DESI-MS datasets
lc_lneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_LNEG.csv", header=1, sep=',')
lc_lpos_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_LPOS.csv", header=1, sep=',')
lc_rneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_RNEG.csv", header=1, sep=',')
lc_rpos_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_RPOS.csv", header=1, sep=',')
lc_hneg_vmet <- read.table("../../Data/LC-MS/VMET_LC-MS_HNEG.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet <- read_csv("../../Data/VMET_Metadata.csv")
cst_metadata_vmet <- read_csv("../../Data/VMET_CSTAssignment.csv")

study_metadata_vmet <- merge(study_metadata_vmet, cst_metadata_vmet, by='Seq_ID')

# Merge with the clinical metadata
lc_lneg_vmet <- merge(study_metadata_vmet, lc_lneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_lpos_vmet <- merge(study_metadata_vmet, lc_lpos_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_rneg_vmet <- merge(study_metadata_vmet, lc_rneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_rpos_vmet <- merge(study_metadata_vmet, lc_rpos_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)
lc_hneg_vmet <- merge(study_metadata_vmet, lc_hneg_vmet, by.x="DESI_ID", by.y="Sample.ID", sort=FALSE)

rm(study_metadata_vmet)
rm(cst_metadata_vmet)
```

Overview of dataset dimensions
```{r}
# VMET LC-MS Lipid Negative Mode
cat(paste("VMET LC-MS Lipid Negative Mode Matrix: ", dim(lc_lneg_vmet)[1], "samples, ", dim(lc_lneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_lneg_first_metabolite_idx_vmet <- 28

# VMET LC-MS Lipid Positive Mode
cat(paste("VMET LC-MS Lipid Positive Mode Matrix: ", dim(lc_lpos_vmet)[1], "samples, ", dim(lc_lpos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_lpos_first_metabolite_idx_vmet <- 28

# VMET LC-MS Reversed Phase Negative Mode
cat(paste("VMET LC-MS Reversed Phase Negative Mode Matrix: ", dim(lc_rneg_vmet)[1], "samples, ", dim(lc_rneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_rneg_first_metabolite_idx_vmet <- 28

# VMET LC-MS Reversed Phase Positive Mode
cat(paste("VMET LC-MS Reversed Phase Positive Mode Matrix: ", dim(lc_rpos_vmet)[1], "samples, ", dim(lc_rpos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_rpos_first_metabolite_idx_vmet <- 28

# VMET LC-MS HILIC Negative Mode
cat(paste("VMET LC-MS HILIC Negative Mode Matrix: ", dim(lc_hneg_vmet)[1], "samples, ", dim(lc_hneg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
lc_hneg_first_metabolite_idx_vmet <- 28
```

## Linear mixed effect modelling of the VMET Dataset
In the next cell we define an utility function which fits a linear mixed model for a single metabolic feature and retrieves the main estimates and p-values. This function will be applied in parallel to all variables in a DESI-MS dataset. This version of the function is tailored for the VMET cohort metadata naming conventions. Model is adjusted for CST, GestationalAge, BMI, Ethnicity and Age, with a random slope and intercept per individual. Contrasts for differences between CSTs, including LDominant (average of I, II, III, V) vs LDepleted CST (IV) are calculated for mid-Gestation (20 weeks), with maternal age = 30, BMI=25 and averaged over all Ethnicity levels. Comparisons between early and late gestation (36 vs 8 weeks) are also calculated, with maternal age=30, BMI=25 and averaged over all Community State Types and Ethnicity levels.
```{r}
fit_lmm_beta_and_pvalues_vmet <- function(dataset, var) {
  
    tryCatch({
      
    currmet <- colnames(dataset)[var]
    
    # LME model were GestationalAge is modelled as a time slope. Random intercept and slope per patient
    rsmodel <- lmer(log(get(currmet) + 1) ~ CST + GestationalAge +  BMI + Ethnicity  + (GestationalAge||SubjectID), 
                    data=dataset,  REML=TRUE)
    
    cst_pairwise_contrasts <- data.frame(emmeans(rsmodel, pairwise~CST, 
                                                  at=list(GestationalAge=20,  BMI=25), adjust='None')$contrast)
    gestage_contrasts <- data.frame(emmeans(rsmodel, revpairwise ~ GestationalAge, 
                                            at=list(GestationalAge=c(8, 36), BMI=25), adjust='None')$contrast)
    
    cst_means <- emmeans(rsmodel, 'CST',  at=list(GestationalAge=20, BMI=25))
    
    lactobacillusDepleted <- data.frame(contrast(cst_means, list(LDepleted=c(-1/4, -1/4, -1/4, 1, -1/4, 0)),  
                                                             at=list(GestationalAge=20, BMI=25), 
                                                             adjust='None'))
    
    pvalues <- c(cst_pairwise_contrasts[, 6], lactobacillusDepleted[, 6], gestage_contrasts[, 6])
    tvalues <- c(cst_pairwise_contrasts[, 5], lactobacillusDepleted[, 5], gestage_contrasts[, 5])
    
    contrast_names <- c(as.character(cst_pairwise_contrasts[ ,1]), as.character(lactobacillusDepleted[, 1]), as.character(gestage_contrasts[, 1]))

    contrast_pval_names <- lapply(contrast_names, FUN=function(x) {paste(x, '_pval')})
    contrast_tval_names <- lapply(contrast_names, FUN=function(x) {paste(x, '_t.ratio')})

    res_frame <- data.frame(cbind(substring(currmet, first=2), t(pvalues), t(tvalues)))
    colnames(res_frame) <- c("Metabolite", contrast_pval_names, contrast_tval_names)
    
    return(res_frame)
        
    }, error = function(err){stop()})
}
```

Run the Linear Mixed Effects modeling in the VMET LC-MS lipid Positive mode VMET data
```{r, warning=F, echo=F, message=F}
lme_lpos_vmet  <- foreach(var=lpos_first_metabolite_idx_vmet:length(lc_lpos_vmet), .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_beta_and_pvalues_vmet(lc_lpos_vmet, var))
    }, error = function(err) {
       # In case of crash fill a row of NA's
       currmet_parsed <- substring(colnames(lc_lpos_vmet)[var], first=2)
       return(c(currmet_parsed, rep(NA,34)))
    })
}


lme_lpos_vmet[2:dim(lme_lpos_vmet )[2]] <- apply(lme_lpos_vmet[, 2:dim(lme_lpos_vmet )[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(lme_lpos_vmet , './LinearMixedModels_LC-MS/VMET_LPOS_LME_CST.csv')
# Apply Benjamini-Hochberg FDR correction at "contrast"/comparison level
lme_lpos_vmet_corrected <- cbind(Metabolite=as.numeric(as.character(lme_lpos_vmet[, 1])), apply(lme_lpos_vmet[, 2:18], 2, p.adjust, method='BH'))

write.csv(lme_lpos_vmet_corrected  , './LinearMixedModels_LC-MS/VMET_LPOS_LME_CST_BH.csv')
```

Run the Linear Mixed Effects modeling in the VMET LC-MS lipid Negative mode VMET data
```{r, warning=F, echo=F, message=F}
lme_lneg_vmet  <- foreach(var=lneg_first_metabolite_idx_vmet:length(lc_lneg_vmet), .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_beta_and_pvalues_vmet(lc_lneg_vmet, var))
    }, error = function(err) {
       # In case of crash fill a row of NA's
       currmet_parsed <- substring(colnames(lc_lneg_vmet)[var], first=2)
       return(c(currmet_parsed, rep(NA,34)))
    })
}


lme_lneg_vmet[2:dim(lme_lneg_vmet )[2]] <- apply(lme_lneg_vmet[, 2:dim(lme_lneg_vmet )[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(lme_lneg_vmet , './LinearMixedModels_LC-MS/VMET_LNEG_LME_CST.csv')
# Apply Benjamini-Hochberg FDR correction at "contrast"/comparison level
lme_lneg_vmet_corrected <- cbind(Metabolite=as.numeric(as.character(lme_lneg_vmet[, 1])), apply(lme_lneg_vmet[, 2:18], 2, p.adjust, method='BH'))

write.csv(lme_lneg_vmet_corrected  , './LinearMixedModels_LC-MS/VMET_LNEG_LME_CST_BH.csv')
```

Run the Linear Mixed Effects modeling in the VMET LC-MS reversed phase Positive mode VMET data
```{r, warning=F, echo=F, message=F}
lme_rpos_vmet  <- foreach(var=rpos_first_metabolite_idx_vmet:length(lc_rpos_vmet), .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_beta_and_pvalues_vmet(lc_rpos_vmet, var))
    }, error = function(err) {
       # In case of crash fill a row of NA's
       currmet_parsed <- substring(colnames(lc_rpos_vmet)[var], first=2)
       return(c(currmet_parsed, rep(NA,34)))
    })
}


lme_rpos_vmet[2:dim(lme_lpos_vmet )[2]] <- apply(lme_rpos_vmet[, 2:dim(lme_rpos_vmet )[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(lme_rpos_vmet , './LinearMixedModels_LC-MS/VMET_RPOS_LME_CST.csv')
# Apply Benjamini-Hochberg FDR correction at "contrast"/comparison level
lme_rpos_vmet_corrected <- cbind(Metabolite=as.numeric(as.character(lme_rpos_vmet[, 1])), apply(lme_rpos_vmet[, 2:18], 2, p.adjust, method='BH'))

write.csv(lme_rpos_vmet_corrected  , './LinearMixedModels_LC-MS/VMET_RPOS_LME_CST_BH.csv')
```

Run the Linear Mixed Effects modeling in the VMET LC-MS lipid Negative mode VMET data
```{r, warning=F, echo=F, message=F}
lme_rneg_vmet  <- foreach(var=rneg_first_metabolite_idx_vmet:length(lc_rneg_vmet), .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_beta_and_pvalues_vmet(lc_rneg_vmet, var))
    }, error = function(err) {
       # In case of crash fill a row of NA's
       currmet_parsed <- substring(colnames(lc_rneg_vmet)[var], first=2)
       return(c(currmet_parsed, rep(NA,34)))
    })
}


lme_rneg_vmet[2:dim(lme_rneg_vmet )[2]] <- apply(lme_rneg_vmet[, 2:dim(lme_rneg_vmet )[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(lme_rneg_vmet , './LinearMixedModels_LC-MS/VMET_RNEG_LME_CST.csv')
# Apply Benjamini-Hochberg FDR correction at "contrast"/comparison level
lme_rneg_vmet_corrected <- cbind(Metabolite=as.numeric(as.character(lme_rneg_vmet[, 1])), apply(lme_rneg_vmet[, 2:18], 2, p.adjust, method='BH'))

write.csv(lme_rneg_vmet_corrected  , './LinearMixedModels_LC-MS/VMET_RNEG_LME_CST_BH.csv')
```

Run the Linear Mixed Effects modeling in the VMET LC-MS lipid Negative mode VMET data
```{r, warning=F, echo=F, message=F}
lme_hneg_vmet  <- foreach(var=hneg_first_metabolite_idx_vmet:length(lc_hneg_vmet), .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_beta_and_pvalues_vmet(lc_hneg_vmet, var))
    }, error = function(err) {
       # In case of crash fill a row of NA's
       currmet_parsed <- substring(colnames(lc_hneg_vmet)[var], first=2)
       return(c(currmet_parsed, rep(NA,34)))
    })
}


lme_hneg_vmet[2:dim(lme_hneg_vmet )[2]] <- apply(lme_hneg_vmet[, 2:dim(lme_hneg_vmet )[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(lme_hneg_vmet , './LinearMixedModels_LC-MS/VMET_HNEG_LME_CST.csv')
# Apply Benjamini-Hochberg FDR correction at "contrast"/comparison level
lme_hneg_vmet_corrected <- cbind(Metabolite=as.numeric(as.character(lme_hneg_vmet[, 1])), apply(lme_hneg_vmet[, 2:18], 2, p.adjust, method='BH'))

write.csv(lme_hneg_vmet_corrected  , './LinearMixedModels_LC-MS/VMET_HNEG_LME_CST_BH.csv')
```

