---
title: "CST Typing by DESI-MS - R2 Linear mixed effect model analysis"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
opts_knit$set(root.dir='./')
```

Load the required packages
```{r, include=FALSE}
library(lme4)
library(ggplot2)
library(doParallel)
library(r2glmm)
library(readr)
library(reshape2)
library(MuMIn)
library(emmeans)

set.seed(774321)
# Set ggplot titles to appear at center
ggplot() + theme(plot.title = element_text(hjust = 'center'))
registerDoParallel(cores=22)
dir.create('./LinearMixedModels_DESI-MS/', recursive=TRUE)
dir.create('./LinearMixedModels_DESI-MS/rSquared', recursive=TRUE)

```

# Data import
Load the VMET2 DESI-MS Positive and Negative mode datasets
```{r, warning=F, echo=F, message=F}
# Load the datasets
desi_neg_vmet2 <- read.table("../../Data/VMET2_DESI-MS_NEG.csv", header=1, sep=',')
desi_pos_vmet2 <- read.table("../../Data/VMET2_DESI-MS_POS.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet2 <- read_csv("../../Data/VMET2_Metadata.csv")
cst_metadata_vmet2 <- read_csv("../../Data/VMET2_CSTAssignment.csv")

study_metadata_vmet2 <- merge(study_metadata_vmet2, cst_metadata_vmet2, by='Seq_ID')

# Merge with the clinical metadata
desi_pos_vmet2 <- merge(study_metadata_vmet2, desi_pos_vmet2, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
desi_neg_vmet2 <- merge(study_metadata_vmet2, desi_neg_vmet2, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
colnames(desi_pos_vmet2)[colnames(desi_pos_vmet2) == 'Seq_ID.x'] <- 'Seq_ID'
colnames(desi_neg_vmet2)[colnames(desi_neg_vmet2) == 'Seq_ID.x'] <- 'Seq_ID'

# Load the data QC results and exclude samples with low spectral quality
desi_neg_vmet2_QC <- read.table("./DESI-MS QC/VMET2_DESI_Negative_QC.csv", header=1, sep=',')
desi_pos_vmet2_QC <- read.table("./DESI-MS QC/VMET2_DESI_Positive_QC.csv", header=1, sep=',')

# Exclude the samples flagged in the DESI-MS Data QC Notebook
neg_vmet2_exclusions <- desi_neg_vmet2[desi_neg_vmet2_QC$Exclude == TRUE, 'DESI_ID']
pos_vmet2_exclusions <- desi_pos_vmet2[desi_pos_vmet2_QC$Exclude == TRUE, 'DESI_ID']
desi_neg_vmet2 <- desi_neg_vmet2[!(desi_neg_vmet2$DESI_ID %in% neg_vmet2_exclusions), ]
desi_pos_vmet2 <- desi_pos_vmet2[!(desi_pos_vmet2$DESI_ID %in% pos_vmet2_exclusions), ]

rm(study_metadata_vmet2)
rm(cst_metadata_vmet2)
rm(desi_neg_vmet2_QC)
rm(desi_pos_vmet2_QC)
rm(neg_vmet2_exclusions)
rm(pos_vmet2_exclusions)
```

Load the VMET DESI-MS Positive and Negative mode datasets
```{r, warning=F, echo=F, message=F}
# Load the DESI-MS datasets
desi_neg_vmet <- read.table("../../Data/VMET_DESI-MS_NEG.csv", header=1, sep=',')
desi_pos_vmet <- read.table("../../Data/VMET_DESI-MS_POS.csv", header=1, sep=',')

# Load the Study Metadata
study_metadata_vmet <- read_csv("../../Data/VMET_Metadata.csv")
cst_metadata_vmet <- read_csv("../../Data/VMET_CSTAssignment.csv")

study_metadata_vmet <- merge(study_metadata_vmet, cst_metadata_vmet, by='Seq_ID')

# Merge with the clinical metadata
desi_pos_vmet <- merge(study_metadata_vmet, desi_pos_vmet, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
desi_neg_vmet <- merge(study_metadata_vmet, desi_neg_vmet, by.x="DESI_ID", by.y="DESI_ID", sort=FALSE)
colnames(desi_pos_vmet)[colnames(desi_pos_vmet) == 'Seq_ID.x'] <- 'Seq_ID'
colnames(desi_neg_vmet)[colnames(desi_neg_vmet) == 'Seq_ID.x'] <- 'Seq_ID'

# Load the data QC results and exclude samples with low spectral quality
desi_neg_vmet_QC <- read.table("./DESI-MS QC/VMET_DESI_Negative_QC.csv", header=1, sep=',')
desi_pos_vmet_QC <- read.table("./DESI-MS QC/VMET_DESI_Positive_QC.csv", header=1, sep=',')

# Exclude the samples flagged in the DESI-MS Data QC Notebook
neg_vmet_exclusions <- desi_neg_vmet[desi_neg_vmet_QC$Exclude == TRUE, 'DESI_ID']
pos_vmet_exclusions <- desi_pos_vmet[desi_pos_vmet_QC$Exclude == TRUE, 'DESI_ID']
desi_neg_vmet <- desi_neg_vmet[!(desi_neg_vmet$DESI_ID %in% neg_vmet_exclusions), ]
desi_pos_vmet <- desi_pos_vmet[!(desi_pos_vmet$DESI_ID %in% pos_vmet_exclusions), ]

rm(study_metadata_vmet)
rm(cst_metadata_vmet)
rm(desi_neg_vmet_QC)
rm(desi_pos_vmet_QC)
rm(neg_vmet_exclusions)
rm(pos_vmet_exclusions)
```

Overview of final dataset dimensions
```{r}
# VMET2 DESI-MS Negative Mode
cat(paste("VMET2 DESI-MS Negative Mode Matrix: ", dim(desi_neg_vmet2)[1], "samples, ", dim(desi_neg_vmet2)[2] - 29, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desineg_first_metabolite_idx_vmet2 <- 29

# VMET2 DESI-MS Positive Mode
cat(paste("VMET2 DESI-MS Positive Mode Matrix: ", dim(desi_pos_vmet2)[1], "samples, ", dim(desi_pos_vmet2)[2] - 29, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desipos_first_metabolite_idx_vmet2 <- 29

# VMET DESI-MS Negative Mode
cat(paste("VMET DESI-MS Negative Mode Matrix: ", dim(desi_neg_vmet)[1], "samples, ", dim(desi_neg_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desineg_first_metabolite_idx_vmet <- 28

# VMET DESI-MS Positive Mode
cat(paste("VMET DESI-MS Positive Mode Matrix: ", dim(desi_pos_vmet)[1], "samples, ", dim(desi_pos_vmet)[2] - 28, "metabolic features", "\n"))
# Assign a variable with the index of the first column corresponding to an MS signal
desipos_first_metabolite_idx_vmet <- 28
```
# Main Linear mixed effect model structure and contrasts
In this notebook, we will define a general mixed effect model formula to model metabolite variability as function of Community State Type, (CST), gestational age, maternal BMI, maternal age, ethnicity, with random effects (slope with gestational age and intercept for baseline differences) to account for between individual variability.

Example LME Model definition and summary
```{r}
# Select a random feature from VMET2
currmet <- colnames(desi_neg_vmet2)[1085]

# metabolite variability explained as a function of CST, GestationalAge, 
# BMI and Ethnicity, with random intercept and slope (GestationalAge) per patient
rsmodel <- lmer(log(get(currmet) + 1) ~ CST + GestationalAge +  BMI + Ethnicity + Age + (0 + GestationalAge|SubjectID) + (1|SubjectID), data=desi_neg_vmet2,  REML=TRUE)

# Default model summmary
summary(rsmodel, ddf='Kenward-Roger')
```

## Linear mixed effect modelling of the VMET2 Dataset
In the next cell we define an utility function which fits a linear mixed model for a single metabolic feature and retrieves the main estimates and p-values. This function will be applied in parallel to all variables in a DESI-MS dataset. This version of the function is tailored for the VMET2 cohort metadata naming conventions. Model is adjusted for CST, GestationalAge, BMI, Ethnicity and Age, with a random slope and intercept per individual. Contrasts for differences between CSTs, including LDominant (average of I, II, III, V) vs LDepleted CST (IV) are calculated for mid-Gestation (20 weeks), with maternal age = 30, BMI=25 and averaged over all Ethnicity levels. Comparisons between early and late gestation (36 vs 8 weeks) are also calculated, with maternal age=30, BMI=25 and averaged over all Community State Types and Ethnicity levels.

Define another helper function to profile the partial R2 per covariate in parallel
```{r}
fit_lmm_r2estimates <- function(dataset, var) {
    tryCatch({
      
      # Get metabolite name
      currmet <- colnames(dataset)[var]
      
      model_variables <- c(currmet, 'CST', 'GestationalAge', 'Age', 'BMI', 'Ethnicity', 'SubjectID')
      complete_idx <- complete.cases(dataset[model_variables])
      
      model_formula <- as.formula(paste0(currmet, " ~ CST + GestationalAge + Age +  BMI + Ethnicity + (GestationalAge||SubjectID)"))
      currentDataset <- dataset[complete_idx, model_variables]
      currentDataset[, 1] <- log(currentDataset[, 1] + 1)
      rsmodel <- lmer(model_formula, data=currentDataset,  REML=TRUE)
      
      r2_partial <- data.frame(r2beta(rsmodel, method='kr'))
      
      rownames(r2_partial) <- r2_partial$Effect
      # Re-order rows so the row binding is coherent in foreach
      r2_partial <- r2_partial[c('Model', 'CST', 'GestationalAge', 'BMI', 'Ethnicity', 'Age'), ]
      mumin_r2 <- r.squaredGLMM(rsmodel)
      r2_cond <- mumin_r2[1]
      r2_marg <- mumin_r2[2]
      icc <- performance::icc(rsmodel, tolerance = 1e-999)
      
      r2_values <- as.numeric(c(r2_partial$Rsq, r2_cond, r2_marg, icc$ICC_adjusted))
      r2_names <- c(as.character(r2_partial$Effect), 'R2Conditional_nsj', 'R2Marginal_nsj', 'Adjusted_ICC')
      r2_names[1] <- 'R2_Model_kr'
      
      pval_values <- as.numeric(c(r2_partial$pval, NA, NA, NA))
      pval_names <- lapply(r2_names, FUN=function(x) {paste0(x, '_pval')})
      
      res_frame <- data.frame(cbind(substring(currmet, first=2), t(r2_values), t(pval_values)))
      colnames(res_frame) <- c("Metabolite", r2_names, pval_names)
      
      return(res_frame)
      
    }, error = function(err){stop()})
}
```

Iterate through the VMET DESI Positive ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_neg_vmet <- foreach(var=desineg_first_metabolite_idx_vmet:dim(desi_neg_vmet)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_neg_vmet, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_neg_vmet)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_neg_vmet[2:dim(r2_estimates_desi_neg_vmet)[2]] <- apply(r2_estimates_desi_neg_vmet[, 2:dim(r2_estimates_desi_neg_vmet)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(r2_estimates_desi_neg_vmet, './LinearMixedModels_DESI-MS/rSquared/VMET_DESI_NEG_LME_R2.csv')
```

Iterate through the VMET DESI Negative ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_pos_vmet <- foreach(var=desipos_first_metabolite_idx_vmet:dim(desi_pos_vmet)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_pos_vmet, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_pos_vmet)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_pos_vmet[2:dim(r2_estimates_desi_pos_vmet)[2]] <- apply(r2_estimates_desi_pos_vmet[, 2:dim(r2_estimates_desi_pos_vmet)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)


write.csv(r2_estimates_desi_pos_vmet, './LinearMixedModels_DESI-MS/rSquared/VMET_DESI_POS_LME_R2.csv')
```

Iterate through the VMET 2 DESI Negative ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_neg_vmet2 <- foreach(var=desineg_first_metabolite_idx_vmet2:dim(desi_neg_vmet2)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_neg_vmet2, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_neg_vmet2)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_neg_vmet2[2:dim(r2_estimates_desi_neg_vmet2)[2]] <- apply(r2_estimates_desi_neg_vmet2[, 2:dim(r2_estimates_desi_neg_vmet2)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)

write.csv(r2_estimates_desi_neg_vmet2, './LinearMixedModels_DESI-MS/rSquared/VMET2_DESI_NEG_LME_R2.csv')
```

Iterate through the VMET 2 DESI Positive ionisation mode dataset and obtain the model R squares and the partial R squares per covariate
```{r}
r2_estimates_desi_pos_vmet2 <- foreach(var=desipos_first_metabolite_idx_vmet2:dim(desi_pos_vmet2)[2], .combine=rbind) %dopar% {
    tryCatch({
        return(fit_lmm_r2estimates(desi_pos_vmet2, var))
    }, error = function(err) {
        # In case of crash fill a row of NA'ss
        currmet_parsed <- substring(colnames(desi_pos_vmet2)[var], first=2)
        return(c(currmet_parsed, rep(NA, 17)))
    })
}

r2_estimates_desi_pos_vmet2[2:dim(r2_estimates_desi_pos_vmet2)[2]] <- apply(r2_estimates_desi_pos_vmet2[, 2:dim(r2_estimates_desi_pos_vmet2)[2]], FUN = function(x) {as.numeric(as.character(x))}, 2)


write.csv(r2_estimates_desi_pos_vmet2, './LinearMixedModels_DESI-MS/rSquared/VMET2_DESI_POS_LME_R2.csv')
```

# Generate a stacked barplot barplot with the R2 measures
```{r}
plotR2 <- function(r2_results, top_n) {
  
  r2_values_set <- r2_results[, c("Metabolite", "CST", "GestationalAge", "Age", "BMI", "Ethnicity", "R2Conditional_nsj", "R2_Model_kr")]
  r2_values_set[, 9] <- r2_values_set[, 7] - rowSums(r2_values_set[, 2:6])
  r2_values_set[r2_values_set[, 9] < 0, 9] <- 0

  colnames(r2_values_set)[9] <- 'R2'

  melted <- melt(r2_values_set, 'Metabolite')

  melted$Metabolite <- round(as.numeric(as.character(melted$Metabolite)), digits=3)
  melted$value <- as.numeric(melted$value)

  covar_to_sort_by <- 'R2Conditional_nsj'
  r2_covar <- subset(melted, variable == covar_to_sort_by)
  r2_covar <- r2_covar[order(-r2_covar$value), ]
  melted$Metabolite <- round(as.numeric(as.character(melted$Metabolite)), digits=3)
  melted$variable <- factor(melted$variable, levels = c('R2Conditional_nsj', 'R2', 'CST', 'GestationalAge', 'Ethnicity', 'BMI', 'Age'))
  top_n_mets <- r2_covar[1:top_n, 1]

  plotDFrame <- melted[melted$Metabolite %in% top_n_mets, ]
  plotDFrame <- subset(plotDFrame, variable != 'R2Conditional_nsj')
  
  r2_plot <- ggplot(plotDFrame, aes(fill=variable, y=as.numeric(value), x=factor(Metabolite, r2_covar$Metabolite[1:top_n]))) +  geom_bar(stat='identity', position='stack', alpha=1) + ylab('R2') + xlab('Metabolite') + theme( axis.text.x = element_blank(),
  axis.ticks.x = element_blank())
  return (r2_plot)
}
```

# Plots for top 100 features (ranked by variance explained in linear model, including the SubjectID) 
```{r}
vmet2_neg_top100 <- plotR2(r2_estimates_desi_neg_vmet2, 100)
vmet_neg_top100 <- plotR2(r2_estimates_desi_neg_vmet, 100)
vmet_pos_top100 <- plotR2(r2_estimates_desi_pos_vmet, 100)
vmet2_pos_top100 <- plotR2(r2_estimates_desi_pos_vmet2, 100)

ggsave('./rSquared/VMET2_DESINEG_100R2.png', vmet2_neg_top100)
ggsave('./rSquared/VMET_DESINEG_100R2.png', vmet_neg_top100)
ggsave('./rSquared/VMET2_DESIPOS_100R2.png', vmet2_pos_top100)
ggsave('./rSquared/VMET_DESIPOS_100R2.png', vmet_pos_top100)
```

# Variance explained for the Markers shown in Figure 1 (B) - main replicated hits of linear mixed modeling analysis.
```{r}
library(xlsx)
ldepl_markers <- read.xlsx('../../Data/Extended Data Table 2 LDOM vs LDEPL.xlsx', sheetIndex=1)

ldepl_markers_neg <- subset(ldepl_markers, DESI.MS.mode =='negative')

ldepl_markers_pos <- subset(ldepl_markers, DESI.MS.mode =='positive')

ldepl_vmet_neg <- r2_estimates_desi_neg_vmet[r2_estimates_desi_neg_vmet$Metabolite %in% ldepl_markers_neg$VMET_m.z.measured, ]
ldepl_vmet2_neg <- r2_estimates_desi_neg_vmet2[r2_estimates_desi_neg_vmet2$Metabolite %in% ldepl_markers_neg$VMET2_m.z.measured, ]

ldepl_vmet_pos <- r2_estimates_desi_pos_vmet[r2_estimates_desi_pos_vmet$Metabolite %in% ldepl_markers_pos$VMET_m.z.measured, ]
ldepl_vmet2_pos <- r2_estimates_desi_pos_vmet2[r2_estimates_desi_pos_vmet2$Metabolite %in% ldepl_markers_pos$VMET2_m.z.measured, ]

formatMz <- function(x) {sapply(x, FUN=function(y){format(round(y, 4), nsmall = 4)})}

immune_vmet2_neg <- r2_estimates_desi_neg_vmet2[formatMz(as.numeric(r2_estimates_desi_neg_vmet2$Metabolite)) %in% formatMz(immune_markers$m.z.measured), ]

vmet_neg_topMarkers <- plotR2(ldepl_vmet_neg, 150)
vmet2_neg_topMarkers <- plotR2(ldepl_vmet2_neg, 150)

vmet_pos_topMarkers <- plotR2(ldepl_vmet_pos, 150)
vmet2_pos_topMarkers <- plotR2(ldepl_vmet2_pos, 150)

ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET_DESINEG_R2_LDomLDeplMarkers.png', vmet_neg_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET2_DESINEG_R2_LDomLDeplMarkers.png', vmet2_neg_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET2_DESIPOS_R2_LDomLDeplMarkers.png', vmet2_pos_topMarkers)
ggsave('./LinearMixedModels_DESI-MS/rSquared/VMET_DESIPOS_R2_LDomLDeplMarkers.png', vmet_pos_topMarkers)
```